\documentclass[a4paper]{report}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{RJournal}
\usepackage{amsmath,amssymb,array}
\usepackage{booktabs}

%% load any required packages here
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

%% do not edit, for illustration only
\sectionhead{Contributed research article}
\volume{XX}
\volnumber{YY}
\year{20ZZ}
\month{AAAA}

%% replace RJtemplate with your article
\begin{article}

% !TeX root = RJwrapper.tex




\title{The mosaic Package}
\author{by Randall Pruim, Nicholas J. Horton and Daniel T. Kaplan}

\maketitle

\abstract{
An abstract of less than 150 words.
}

Introductory section which may include references in parentheses
\citep{R}, or cite a reference such as \citet{R} in the text.

\section{Project MOSAIC}

\iffalse
Project MOSAIC is an NSF-funded initiative to provide a broader approach to
quantitative studies that provides better support for work in science and
technology. The focus of the project has been to tie together better diverse aspects
of quantitative work that students in science, technology, and engineering will
need in their professional lives, but which are usually taught in isolation, if at all.

The name MOSAIC reflects a vision of 
quantitative education where the basic components come together to form a complete 
and compelling picture.
The name also loosely reflects the first letters -- M, S, C, C -- of four important
components of a quantitative education:
\begin{description}
\item{Modeling.} The ability to create, manipulate and investigate useful and informative mathematical representations of a real-world situations.
\item{Statistics.}
	The analysis of variability that draws on our ability to quantify uncertainty and to draw logical inferences from observations and experiment.
\item{Computation.} The capacity to think algorithmically, to manage data on large scales, to visualize and interact with models, and to automate tasks for efficiency, accuracy, and reproducibility.
\item{Calculus.}
The traditional mathematical entry point for college and university students and a subject that still has the potential to provide important insights to today’s students.
\end{description}
\fi

The \pkg{mosaic} package originated in early attempts by each of the authors to 
ease new users into using R, primarily in the context of 
undergraduate statistics courses, and, in one case, also in calculus.

\section{Expanding the formula interface}

\subsection{Less volume, more creativity}

One of the guiding principles behind the development of the \CRANpkg{mosaic} package
has been ``Less volume, more creativity".  Beginners are easily overwhelmed by the 
scope of R and its many packages:  Often there are multiple ways to accomplish the 
same task, and authors of the many packages are not required to follow any particular
style guidelines.  Early on in the development of \CRANpkg{mosaic}, we decided to
reduce the number of code templates that users would need to know to as few as possible,
while still providing them with substantial power to be creative within the templates provided.

Our most important template makes use what we call the ``formula-interface'' and is modeled
after \code{lm()} and the plotting functions in \CRANpkg{lattice}.  We present the template as

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
goal(y ~ x | z, data = mydata)
\end{verbatim}
\end{kframe}
\end{knitrout}
or even more simply as
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
goal(formula, data = mydata)
\end{verbatim}
\end{kframe}
\end{knitrout}
This emphasizes that functions in R are typically named after the object they produce (the goal) and
that a wide variety of goals can be achieved using a single template.  Having learned the formula
interface to make \CRANpkg{lattice} plots, new users are well prepared for modeling with
\code{lm()} and \code{glm()} when the time comes. 

\subsection{Adding the formula interface to existing functions}

Although this formula interface works well for graphical summaries of data, the
same is not true of the standard numerical summary functions (e.g.,
\code{mean()}, \code{median()}, \code{sd()}, etc.).  One of our initial
difficult decisions was whether to add such a formula interface to these
standard functions (by masking them) or create new functions.  We decided on the former 
option since the standard names are natural and familiar and having two sets of functions for 
essentially the same task would be unnecessarily complicated for new users.  The \CRANpkg{mosaic} 
versions of these functions can still be used as in \CRANpkg{stats} and \CRANpkg{base} so that
existing code should not be affected.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
mean(HELPrct$age)
[1] 35.65342
mean(~age, data = HELPrct)  # same result as above 
[1] 35.65342
mean(age ~ sex, data = HELPrct)  # additional features are possible
  female     male 
36.25234 35.46821 
\end{verbatim}
\end{kframe}
\end{knitrout}
There are, of course, other ways to achieve this last result, including the use of 
\code{aggregate()}, or \code{summary()} from \CRANpkg{Hmisc}, or the functions in 
\CRANpkg{plyr}.   But each of these solutions would require learning additional coding 
templates.  In our approach, the following (and many more) are all identical except for 
the specification of the desired result via the function name.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
   bwplot( age ~ sex, data=HELPrct )
     mean( age ~ sex, data=HELPrct )
       sd( age ~ sex, data=HELPrct )
       lm( age ~ sex, data=HELPrct )
   t.test( age ~ sex, data=HELPrct )   # formula interface exists in stats
\end{verbatim}
\end{kframe}
\end{knitrout}
Similarly, for one-variable situations we have 
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
      mean( ~ age, data=HELPrct )
        sd( ~ age, data=HELPrct )
 histogram( ~ age, data=HELPrct )
    t.test( ~ age, data=HELPrct )       # formula interface added in mosaic
binom.test( ~ sex, data=HELPrct )       # formula interface added in mosaic
 prop.test( ~ sex, data=HELPrct )       # formula interface added in mosaic
\end{verbatim}
\end{kframe}
\end{knitrout}
And for plots and numerical summaries, we can add covariates into the mix as well:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
     mean( ~ age | sex, data=HELPrct )
       sd( ~ age | sex, data=HELPrct )
histogram( ~ age | sex, data=HELPrct )
   t.test( ~ age | sex, data=HELPrct ) # formula interface added in mosaic (TO DO!!)
\end{verbatim}
\end{kframe}
\end{knitrout}

\section{Some things that we didn't change}

The formula interface is powerful and flexible.  Nevertheless, there are some sticky issues that we 
have chosen not to address.

		\subsection{The geometry of \CRANpkg{lattice} formlas}

		In \code{lm()}, \code{glm()} and the numerical summary functions, \code{y ~ x} can
		typically be read as 
		``\code{y} depends on \code{x}'' or
		``\code{y} is modeled by \code{x}''.  From this perspective,
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
histogram(age ~ sex, data = HELPrct)
\end{verbatim}
\end{kframe}
\end{knitrout}
would be perfectly reasonable.  But \CRANpkg{lattice}, for the most part, 
uses the \code{y} and \code{x} slots to indicate which axis of the plot is used 
for which variable. So the correct \CRANpkg{lattice} command for the desired
pair of histograms is 
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
histogram(~age | sex, data = HELPrct)
\end{verbatim}
\end{kframe}
\end{knitrout}
and for overlaid density plots it is 
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
histogram(~age, groups = sex, data = HELPrct)
\end{verbatim}
\end{kframe}
\end{knitrout}
While this can produce some initial challenges for users, clearly explaining the roles 
of the components of the variables for plotting, for numerical summaries, and for model 
fitting helps demystify the situation.  Furthermore, in several functions we promote formulas
of the form \code{~ x | z} to \code{x ~ z} to allow either format.  Similarly, the \code{groups}
argument can often be used, for example
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
mean(~age, groups = sex, data = HELPrct)
  female     male 
36.25234 35.46821 
\end{verbatim}
\end{kframe}
\end{knitrout}

\subsection{Handling missing data}


\subsection{Other stuff as we think of it}


\section{Extracting information}

Modeled on functions like \code{resid()}, a number of additional functions have
been added to \CRANpkg{mosaic} to facilitate extracting information from more
complicated objects.  Some examples include
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
confint(t.test(~age, data = HELPrct))  # works for 'htest' objects
mean of x     lower     upper     level 
 35.65342  34.94150  36.36534   0.95000 
pval(t.test(age ~ sex, data = HELPrct))  # works for 'htest' objects
  p.value 
0.3537401 
stat(t.test(age ~ sex, data = HELPrct))  # works for 'htest' objects
        t 
0.9297644 
r.squared(lm(age ~ sex, data = HELPrct))
\end{verbatim}


{\ttfamily\noindent\color{warningcolor}{Warning: 'r.squared' is deprecated.\\Use 'rsquared' instead.\\See help("{}Deprecated"{})}}\begin{verbatim}
[1] 0.001870074
\end{verbatim}
\end{kframe}
\end{knitrout}

\section{Simplifying randomization}

\subsection{Using do() for repetition}

\subsection{sample(), resample(), and shuffle()}

\subsection{do() and confint()}

\subsection{statTally()}


\section{Calculus in R}

\section{Some additional bells and whistles}


\section{Summary}

This file is only a basic article template. For full details of \emph{The R
Journal} style and information on how to prepare your article for submission,
see the
\href{http://journal.r-project.org/latex/RJauthorguide.pdf}{Instructions for
Authors}.

\bibliography{pruim-horton-kaplan}

\address{Randall Prim \\
  Calvin College\\
  3201 Burton St SE\\
  Grand Rapids, MI  49546\\
  USA}
\email{rpruim@calvin.edu}

\address{Author Two\\
  Affiliation\\
  Address\\
  Country}
\email{author2@work}

\address{Author Three\\
  Affiliation\\
  Address\\
  Country}
\email{author3@work}
%  \input{pruim-horton-kaplan}
\end{article}

\end{document}
