---
title: "ggformula/lattice conversion examples"
author: "Nicholas Horton (nhorton@amherst.edu)"
date: "July 16, 2017"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
    pandoc_args: [
      "-V", "classoption=twocolumn"
    ]
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---
```{r, setup, include=FALSE}
require(mosaic)   # Load additional packages here 

# Some customization.  You can alter or delete as desired (if you know what you are doing).
trellis.par.set(theme=theme.mosaic()) # change default color scheme for lattice
knitr::opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
```

## Introduction
This document is intended to help users of the `mosaic` package migrate their `lattice` package graphics to `ggformula`.  The mosaic package provides a simplified and systematic introduction to the core functionality related to descriptive statistics, visualization, modeling, and simulation-based inference required in first and second courses in statistics.  Originally, the `mosaic` package used `lattice` graphics but now support is also available for the `ggformula` system. 


## Histograms (ggformula)

```{r}
library(mosaic)  # also loads ggformula
gf_histogram(~ age, data=HELPrct)
```

## Histogram options (ggformula)

```{r}
gf_histogram(~ age, binwidth=5, 
  data=HELPrct)   # or age ~ 1
```

\newpage

## References 

More information about `ggformula` can be found at https://github.com/ProjectMOSAIC/ggformula.

More information regarding Project MOSAIC (Kaplan, Pruim, and Horton) can be found at http://www.mosaic-web.org.  Further information regarding the `mosaic` package can be found at https://github.com/ProjectMOSAIC/mosaic and https://journal.r-project.org/archive/2017/RJ-2017-02.


## Histograms (lattice)

```{r}
library(mosaic) # also loads lattice
histogram(~ age, data=HELPrct)
```

## Histogram options (lattice)

```{r}
histogram(~ age, width=5, data=HELPrct)
```




\newpage

## Density plots (ggformula)

```{r, message=FALSE}
gf_dens(~ age, data=HELPrct)
```

## Overlaid density plots (ggformula)

```{r, message=FALSE}
gf_dens(~ age,  color = ~ sex, data=HELPrct)
```

\newpage

## Density plots (lattice)

```{r, message=FALSE}
densityplot(~ age, data=HELPrct)
```

## Overlaid density plots (lattice)

```{r}
densityplot(~ age, groups=sex, data=HELPrct)
```

\newpage

## Side by side box plots (ggformula)

```{r, message=FALSE}
gf_boxplot(age ~ sex, data=HELPrct)
```

## Faceted side by side box plots (ggformula)

```{r, message=FALSE}
gf_boxplot(age ~ sex | homeless, data=HELPrct)
```


\newpage

## Side by side box plots (lattice)



## Side by side plots (lattice)

```{r, message=FALSE}
bwplot(age ~ sex, data=HELPrct)
```

## Faceted side by side plots (lattice)

```{r, message=FALSE}
bwplot(age ~ sex | homeless, data=HELPrct)
```

\newpage

## Scatterplot (lattice)

```{r, message=FALSE}
xyplot(cesd ~ age, data=HELPrct)
```



## Overlaid scatterplot with linear fit (ggformula)

```{r, message=FALSE}
gf_point(cesd ~ age, color = ~ sex, 
         data=HELPrct) %>%
  gf_lm()
``````

## Faceted scatterplot with smooth fit (ggformula)

```{r, message=FALSE}
gf_point(cesd ~ age | sex, 
         data=HELPrct) %>%
  gf_smooth()
``````

\newpage

## Overlaid scatterplot with linear fit (lattice)

```{r, message=FALSE}
xyplot(age ~ cesd, groups=sex, type=c("p", "r"), 
       auto.key=TRUE, data=HELPrct)
```

## Faceted scatterplot with smooth fit (lattice)

```{r, message=FALSE}
xyplot(age ~ cesd | sex, type=c("p", "smooth"), 
       auto.key=TRUE, data=HELPrct)
```


\newpage

## More options for scatterplot with linear fit (ggformula)

```{r}
gf_point(cesd ~ age, color = ~ sex, 
         data=HELPrct) %>%
  gf_lm(se=TRUE) %>% 
  gf_theme(legend.position = "top") %>% 
  gf_labs(title = "This is my plot", 
    x="age (in years)", 
    y="CES-D measure of\ndepressive symptoms")
```

\newpage

## More options for scatterplot with linear fit (lattice)

```{r, message=FALSE}
xyplot(age ~ cesd, groups=sex, 
       type=c("p", "r"), 
       auto.key=TRUE, 
       main = "This is my plot", 
       xlab="age (in years)", 
       ylab="CES-D measure of\ndepressive symptoms",
       data=HELPrct)
```







