---
title: "Examples for Issues"
author: "R Pruim"
date: "May 16, 2015"
output: html_document
---

```{r include=FALSE}
require(mosaic)
```

## Identifying lazy data and success argument in binom.test() and prop.test() output

```{r}
# using ggplot2::midwest as an example
binom.test(~category, data=midwest)
prop.test(~category, data=midwest)
```

Note:  For lazy loaded data, using `deparse(lazyeval::lazy(data)$expr)` 
doesn't contain what we want, so I'm reverting to using `substitute()` here.

# Error messages that could be improved

## favstats error message (#458)

This error message was previously cryptic.  Better now.
```{r error=TRUE, purl=FALSE}
require(mosaicData)
favstats(~sex, data=HELPrct)
```

## CIAdata()

Adding in ability to select multiple names and get a merged data set.
```{r}
World <- CIAdata(c("pop", "area", "life", "fert"))
head(World)
```


## Issue 736: xpt() bugs

This was throwing an error because of missing `ncp`:

```{r}
suppressPackageStartupMessages(library(mosaic, quietly = TRUE))
xpt(.67, df = 49)
```

Trouble caused because there is no default for `ncp` in `pt()`  Indeed

> Supplying ncp = 0 uses the algorithm for the non-central distribution, which is not the same algorithm used if ncp is omitted. This is to give consistent behaviour in extreme cases with values of ncp very near zero.

Fixed with explicit test for missing.

## Issue 737

`...` was missing in a couple of function and so `xlim` (and other things) were
not being passed along.

```{r}
library(mosaic) #latest vers.
xpnorm(3, xlim = c(2, 4))
xpt(41.944,df=4, ncp=0, xlim=c(41,43))
xpchisq(41.944,df=4, ncp=0, xlim=c(41,43))
xpf(41.944,df1=4,df2=3, ncp=0, xlim=c(41,43))
# In previous mosaic vers. all the above "zooms" were working very well after adding
# the "default" ncp=0 - in the latest vers. 1.5.0
# xpchisq stopped with the error message:
# Error in xpchisq(41.944, df = 4, ncp = 0, xlim = c(41, 43)) :
# unused argument (xlim = c(41, 43))
```
